---
title: "Course Project 1: Step Analysis"
output: html_document
date: "2024-05-09"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressMessages({
        suppressWarnings(
        {
                library(dplyr)
                library(ggplot2)
                library(cowplot)
        })
})
```

## Part 1: Load and preprocess the data. 

### 1. Load data
```{r}
data <- read.csv('activity.csv')
head(data)
```

## Part 2: What is mean total number of steps taken per day?

### A. Calculate the total number of steps taken per day.
```{r cars}
sum_data <- data %>% group_by(date) %>% summarize(Sum_Steps = 
                                                          sum(steps, na.rm=TRUE))
```

### B. Make a histogram of the total number of steps taken each day.
```{r}
hist(sum_data$Sum_Steps,
     main="Total Number of Steps Taken Each Day",
     xlab="Steps Taken",
     ylab='Frequency of Steps Taken',
     ylim=c(0, 30),
     col="darkmagenta"
     )
```

### C. Calculate the mean and median of the total of steps taken per day. 
```{r}
mean <- mean(sum_data$Sum_Steps, na.rm = TRUE)
median <- median(sum_data$Sum_Steps, na.rm = TRUE)
print(paste('Mean of total of steps taken per day is', mean))
print(paste('Median of total of steps taken per day is', median))
```

## Part 3: What is the average daily activity pattern?

### A. Make a time series plot of the 5-minute interval (x-axis) and <br> 
### the average number of steps taken, averaged across all days.
```{r}
interval_data <- data %>% group_by(interval) %>% 
        summarize(Mean_Interval = mean(steps, na.rm=TRUE))

p <- ggplot(interval_data, aes(x=interval, y=Mean_Interval)) + 
            geom_line(color="turquoise4") + theme_minimal() + 
            labs(x="Five-Minute Intervals", 
                 y="Mean Number of Steps", 
                 title="Mean Number of Steps for Each Five-Minute Interval") 
print(p)
```

### B. Which 5-minute interval, on average across all the days in the data set, <br> 
### contains the maximum number of steps?
```{r}
max <- max(interval_data$Mean_Interval, na.rm = TRUE)
print(paste('Max of mean of total of steps taken each five-minute interval is'
            , max))
```

## Part 4: Imputing missing values

### A. Calculate and report the total number of missing values in the data set.
```{r}
frac_missing <- sum(is.na(data$steps))  / length(data$steps)
print(paste('Proportion of missing values is', frac_missing))
```

### B. Devise a strategy for filling in all of the missing values in the data set.

Missing values don't seem to be randomly distributed,
and without more information about why they are missing, <br>
removing entries with missing values seems to be the best approach.
So my approach is to locate rows with missing values, and delete them. 

### C. Create a new data set with the missing data filled in.
```{r}
new_data <- data[complete.cases(data), ]
head(new_data)
```

### D. Make a histogram of the total number of steps taken each day.

#### Calculate total number of steps 
```{r}
new_sum_data <- new_data %>% group_by(date) %>% 
                summarize(Sum_Steps = sum(steps, na.rm=TRUE))
head(new_sum_data)
```

#### Plot histogram 
```{r}
hist(new_sum_data$Sum_Steps,
     main="Total Number of Steps Taken\nEach Day",
     xlab="Steps Taken",
     ylab='Frequency of Steps Taken',
     ylim=c(0, 30),
     col="royalblue4"
)
```

## E. Calculate and report the mean and median total number of steps taken per day.
```{r}
new_mean <- mean(new_sum_data$Sum_Steps, na.rm = TRUE)
new_median <- median(new_sum_data$Sum_Steps, na.rm = TRUE)
print(paste('Mean of total of steps taken per day is', new_mean))
print(paste('Median of total of steps taken per day is', new_median))
```

## F. What is the impact of imputing missing data on the estimates? 
```{r}
perc_mean_diff <- (mean - new_mean) / mean(mean, new_mean) * 100 
print(paste('The percent difference between the first mean and the second is ', 
            perc_mean_diff, '%', sep=''))
perc_median_diff <- (median - new_median) / mean(median, new_median) * 100 
print(paste('The percent difference between the first median and the second is ',
            perc_median_diff, '%', sep=''))

```

After dropping the missing values, the mean dropped about 15%,
and the median dropped by about 3.5%.

## Part 5: Are there differences in activity between weekdays and weekends?

### A. Create a new factor variable for “weekday” and “weekend”.

#### Convert date column to POSIXlt class 
```{r}
new_data$day <- strftime(new_data$date, "%A")
```

#### Find weekend values 
```{r}
weekend <- c('Sunday', 'Saturday')
```

#### Create mask to find weekend days 
```{r}
mask <- new_data$day %in% weekend
```

#### Create factor column 'day_category'
```{r}
new_data$day_category <- factor(mask, 
                                levels=c(FALSE, TRUE),
                                labels = c("weekday", "weekend"))
head(new_data)
```

### B. Make a panel plot containing a time series plot of the 5-minute interval <br> 
### and average number of steps taken, averaged across all the days. 
```{r}
new_interval_data <- new_data %>% group_by(day_category, interval) %>% 
        summarize(Mean_Interval = mean(steps))
```

#### Separate data into weekend and weekday data frames 
```{r}
weekend_data <- new_interval_data[new_interval_data$day_category %in% c('weekend'), ]
weekday_data <- new_interval_data[new_interval_data$day_category %in% c('weekday'), ]
```

#### Create panel plot

##### Create weekday data plot
```{r}
p1 <- ggplot(weekday_data, aes(x=interval, y=Mean_Interval)) + 
        geom_line(color="deepskyblue4") + theme_minimal() + 
        labs(x="Five-Minute Intervals", 
             y="Mean Number of Steps", 
             title="Mean Number of Steps for Each Five-Minute Interval on Weekdays") 
```

##### Create weekend data plot
```{r}
p2 <- ggplot(weekend_data, aes(x=interval, y=Mean_Interval)) + 
        geom_line(color="darkolivegreen4") + theme_minimal() + 
        labs(x="Five-Minute Intervals", 
             y="Mean Number of Steps", 
             title="Mean Number of Steps for Each Five-Minute Interval on Weekends") 
```

##### Plot on panel 
```{r}
plot_grid(p1, p2, ncol = 1, nrow = 2)
```

